Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-07 brms_meeg.qmd
updating the simulation code (now faster and more parallelised)

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    dev = "cairo_pdf"
    )

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography (MEG) and electroencephalography (EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether—and when—neural responses differ across experimental conditions or participant groups.

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @pedersen_hierarchical_2019; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also @maris2011 and @rosenblatt2018 (history of cluster-based approaches)... Clusters failures [@eklund2016]...

@@

## Previous work

@@

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $\mathbf{A}_{i}$ is the $i$th row of a known parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms (to be estimated), $f_{j}$ is a smooth function of covariate $x_{j}$ (to be estimated as well). The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions.

@@

@@

Focusing on identifying onset and offset of effects (as assessed by ERP differences or decoding performance)... Assessing the performance of a model-based approach (i.e., Bayesian GAMMs) to conventional methods (multiplicity corrections or cluster-based permutation)...

@@

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

```{r fig-eeg, echo = FALSE, fig.width = 9, fig.cap = "Averaged (mean) simulated EEG activity in two conditions with 50 trials each, for a group of 20 participants. The error band represents the mean plus/minus 1 standard error of the mean."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

# importing helper functions
source("code/functions.R")

# to use with the eeg_noise function
meanpower <- unlist(read.table("code/meanpower.txt") )

@@

# removing raw_df
if (exists("raw_df") ) rm(raw_df)

@@

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.

```{r gam, message = FALSE}
# averaging across participants
ppt_df <- raw_df %>%
    group_by(participant, condition, time) %>%
    summarise(eeg = mean(eeg) ) %>%
    ungroup()

@@

```{r gamm, echo = FALSE, eval = FALSE, message = FALSE}
# defining a contrast for condition
contrasts(raw_df$condition) <- c(-0.5, 0.5)

@@

## Posterior probability of difference above 0

We then plot the posterior predictions together with the posterior estimate of the slope for `condition` at each timestep (@fig-plot-post-slope).

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior estimate of the ERP in each condition (left) or directly for the difference of ERPs (right) according to the GAM."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL) {

@@

We then compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-ratio, left), with $\epsilon := 0.05$, which can be interpreted as the smallest effect size of interest. We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio, right).

```{r fig-post-prob-test, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3, n_draws = 1e3) {

@@

```{r fig-post-prob-ratio, echo = FALSE, out.width = "100%", fig.width = 9, fig.asp = 0.5, fig.cap = "Left: Posterior probability of the ERP difference (slope) being above 0 according to the GAM. Right: Ratio of posterior probability according to the GAM (on a log10 scale). Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

## Multilevel modelling using ERP summary statistics

Next we fit a multilevel version of the GAM. Although it is possible to fit a GAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD, similar to what is done in meta-analysis...

@@

We then computed the difference between the true and estimated onset/offset of the ERP difference ($\text{error}:=|\hat{\theta}-\theta|$), according to various `k` (GAM basis dimension) and `threshold` values. Remember that the signal is generated from a truncated Gaussian with an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the hierarchical GAM can *exactly* recover the true onset and offset values, given some reasonable choice of `k` and `threshold` values (e.g., a threshold of 20). We provide more detailed recommendations on how to set `k` in @sec-basis.

```{r fig-onset-error, eval = TRUE, echo = FALSE, results = "hide", dpi = 300, fig.width = 9, fig.asp = 0.5, fig.cap = "Error function of onset (left) and offset (right) estimation according to various basis dimension and threshold values (according to the multilevel GAM)."}
# defining values of k and threshold to test
# threshold_values <- seq(from = 1, to = 50, by = 1)
# kvalues <- c(5, 10, 15, 20, 25, 30, 35, 40)

@@

# number of posterior samples to use
# n_post_samples <- 1e3

@@

# initialising results dataframe
# results <- crossing(kvalue = kvalues, threshold = threshold_values) %>%
# results <- crossing(kvalue = kvalues, sim_id = 1:nsims, threshold = threshold_values) %>%
#     mutate(
#         estimated_onset = NA,
#         estimated_offset = NA,
#         error_onset = NA,
#         error_offset = NA
#         )

# looping over different values of kvalue and threshold
# for (i in seq_len(nrow(results) ) ) {
# 
#     # retrieving current kvalue and threshold
#     kvalue <- results$kvalue[i]
#     threshold <- results$threshold[i]
# 
#     # printing progress
#     cat(
#         "\nAssessing combination:", i, "out of", nrow(results), "combinations...\n",
#         "k-value:", kvalue, "\n", "threshold:", threshold, "\n"
#         )
# 
#     # when we assess a new k-value
#     if (threshold == 1) {
#         
#         # generate some data
#         raw_df <- generate_data(
#             n_trials = n_trials, n_ppt = n_ppt, outvar = outvar,
#             srate = srate, ronset = ronset
#             )
#         
#         # sanity check
#         head(raw_df)
#         
#         # averaging across participants
#         summary_df <- raw_df %>%
#             summarise(
#                 eeg_mean = mean(eeg),
#                 eeg_sd = sd(eeg),
#                 .by = c(participant, condition, time)
#                 )
#         
#         # defining a contrast for condition
#         contrasts(summary_df$condition) <- c(-0.5, 0.5)
# 
#         # fitting the GAM
#         error_model <- brm(
#             # using by-participant SD of EEG across trials
#             eeg_mean | se(eeg_sd) ~
#                 condition + s(time, bs = "cr", k = kvalue, by = condition) +
#                 (1 | participant),
#             data = summary_df,
#             family = gaussian(),
#             warmup = 2000,
#             iter = 5000,
#             chains = 8,
#             cores = 8
#             )
# 
#     }
# 
#     prob_y_above_0 <- error_model$data %>%
#         add_epred_draws(object = error_model, ndraws = n_post_samples) %>%
#         # add_epred_draws(object = meta_gam) %>%
#         data.frame() %>%
#         dplyr::select(participant, time, condition, .epred, .draw) %>%
#         # dplyr::select(time, condition, .epred, .draw) %>%
#         pivot_wider(names_from = condition, values_from = .epred) %>%
#         mutate(epred_diff = cond2 - cond1) %>%
#         # computing mean posterior probability at the participant level
#         # group_by(participant, time) %>%
#         # summarise(m = mean(epred_diff > 0) ) %>%
#         # ungroup() %>%
#         # computing mean posterior probability at the group level
#         group_by(time) %>%
#         # summarise(m = mean(m) ) %>%
#         summarise(m = mean(epred_diff > 0) ) %>%
#         mutate(prob_ratio = m / (1 - m) ) %>%
#         ungroup()
# 
#     # finding onset, offset, and peak for the current threshold
#     exceeding_times <- prob_y_above_0 %>%
#         dplyr::filter(prob_ratio > threshold) %>%
#         summarise(
#             cluster_onset = min(time, na.rm = TRUE),
#             cluster_offset = max(time, na.rm = TRUE)
#             ) %>%
#         mutate(true_onset = true_onset, true_offset = true_offset)
# 
#     # storing the results in the dataframe
#     if (nrow(exceeding_times) > 0) {
# 
#         results$estimated_onset[i] <- exceeding_times$cluster_onset
#         results$estimated_offset[i] <- exceeding_times$cluster_offset
# 
#         # computing errors
#         results$error_onset[i] <- abs(
#             exceeding_times$cluster_onset - exceeding_times$true_onset
#             )
#         results$error_offset[i] <- abs(
#             exceeding_times$cluster_offset - exceeding_times$true_offset
#             )
# 
#     }
# 
# }

@@

# loading the simulation results
results <- readRDS(file = "results/meta_gam_error_properties.rds")

results %>%
    group_by(kvalue, threshold) %>%
    summarise(across(estimated_onset:error_offset, mean) ) %>%
    ungroup() %>%
    pivot_longer(cols = error_onset:error_offset) %>%
    mutate(name = factor(x = name, levels = c("error_onset", "error_offset") ) ) %>%
    ggplot(
        aes(
            x = as.factor(kvalue),
            y = as.factor(threshold),
            fill = value
            )
        ) +
    # geom_raster(interpolate = FALSE) +
    geom_tile() +
    # geom_point(
    #     data = results %>%
    #         group_by(kvalue, threshold) %>%
    #         summarise(across(estimated_onset:error_offset, mean) ) %>%
    #         ungroup() %>%
    #         na.omit() %>%
    #         filter(error_onset == min(error_onset) ),
    #     # aes(x = kvalue, y = threshold),
    #     aes(x = kvalue, y = value),
    #     color = "orangered", size = 4, shape = 4,
    #     # show.legend = FALSE
    #     show.legend = TRUE
    #     ) +
    # scale_x_continuous(expand = c(0, 0) ) +
    # scale_y_continuous(expand = c(0, 0) ) +
    facet_wrap(~name) +
    scale_x_discrete(expand = c(0, 0) ) +
    scale_y_discrete(expand = c(0, 0), breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50) ) +
    # scale_y_continuous(n.breaks=10) +
    scale_fill_gradientn(colors = rev(met.brewer("Hokusai1") ) ) +
    # scale_fill_viridis_c(direction = 1) +
    labs(
        x = "GAM basis dimension (k)",
        y = "Threshold",
        fill = "Error (s)"
        )

@@

# displaying best parameters values (according to both onset and offset error)
# results %>%
#     group_by(kvalue, threshold) %>%
#     summarise(across(estimated_onset:error_offset, mean) ) %>%
#     ungroup() %>%
#     arrange(error_onset, error_offset) %>%
#     data.frame() %>%
#     head(40)
```

@@

To provide a more exhaustive assessment of the performance of the proposed approach, we compared the ability of the GAMM to correctly estimate the onset and offset of the ERP difference to widely-used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.05$). We then followed the same approach but after applying different forms of multiplicity correction to the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the family-wise error rate (FWER) [i.e., Holm–Bonferroni method, @holm1979], and two cluster-based permutation methods [permutation with a single cluster-forming threshold and threshold-free cluster enhancement, TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. @fig-corrections illustrates the onsets and offsets estimated by each method on a single simulated dataset and shows that all methods over-estimate the true onset and under-estimate the true offset.

<!--

@@

-->

```{r fig-corrections, echo = FALSE, results = "hide", fig.width = 8, fig.asp = 0.5, fig.cap = "Exemplary timecourse of squared t-values with true onset and offset (vertical black dashed lines) and onsets/offsets identified using the raw p-values, the corrected p-values (BH95, BY01, Holm), the cluster-based methods (Cluster-mass, TFCE), or using the binary segmentation method (Change point)."}
# defining an arbitrary alpha threshold
aath <- 0.05

@@

The various methods were assessed using the median error and variance of onset/offset estimates computed on 10.000 simulated datasets...

@@

We assessed the reliability of the proposed approach using a form of permutation-based split-half reliability [as for instance in @rosenblatt2018], which consisted of the following steps. First, we created 1000 split halves of the data (i.e., with half the participants in the original data, that is, 16 participants). For each split, we estimated the onset/offset using all methods described previously. Third, we summarised the distribution of onset/offset estimates using the median "error" (i.e., difference between the split estimate and the estimate obtained using the full dataset) and the variance across splits.

<!--

@@

## Simulation results (bias and variance)

>>>>

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    # dev = "cairo_pdf",
    dev = "png",
    dpi = 300
    )

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography (MEG) and electroencephalography (EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether and when neural responses differ across experimental conditions or participant groups.

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also @maris2011 and @rosenblatt2018 (history of cluster-based approaches)... Clusters failures [@eklund2016]...

@@

## Related work

@@

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $\mathbf{A}_{i}$ is the $i$th row of a known parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms (to be estimated), $f_{j}$ is a smooth function of covariate $x_{j}$ (to be estimated as well). The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of $K$ simpler, basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised.

@@

Introduction to multilevel GAMs [@pedersen_hierarchical_2019]...

@@

Given the previously reported limitations of conventional methods to precisely identify the onset and offset of M/EEG effects (e.g., ERPs, decoding performance), we aimed to provide a model-based approach for estimating the onset and offset of such effects. To achieve this, we leveraged Bayesian generalised additive multilevel models (BGAMMs) fitted via the `brms R` package and compared the performance of this approach to conventional methods on both simulated and actual M/EEG data.

@@

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see code in the online supplementary materials and details in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

```{r fig-eeg, echo = FALSE, fig.width = 9, fig.cap = "Mean simulated EEG activity in two conditions with 50 trials each, for a group of 20 participants. The error band represents the mean +/- 1 standard error of the mean."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

# importing helper functions
# source("code/functions.R")

# EEG power information to use with the eeg_noise() function
meanpower <- unlist(read.table("code/meanpower.txt") )

@@

# removing raw_df if it already exists
if (exists("raw_df") ) rm(raw_df)

@@

```{r fig-erp, eval = FALSE, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$ [@gabry2019; @gelman2020]. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.

```{r gam, results = "hide"}
# averaging across participants
ppt_df <- raw_df %>%
    group_by(participant, condition, time) %>%
    summarise(eeg = mean(eeg) ) %>%
    ungroup()

@@

```{r gamm, echo = FALSE, eval = FALSE, results = "hide"}
# defining a contrast for condition
contrasts(raw_df$condition) <- c(-0.5, 0.5)

@@

We depict the posterior predictions together with the posterior estimate of the slope for `condition` at each timestep (@fig-plot-post-slope).

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior estimate of the EEG activity in each condition (left) and posterior estimate of the difference in EEG activity (right) according to the GAM."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL) {

@@

We then compute the posterior probability of the slope for `condition` being above $0$ (@fig-post-prob-ratio, left). We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio, right). A ratio of 10 means that the probability of the difference being above 0 is 10 times higher than the probability of the difference not being above 0, given the data, the priors, and other model's assumptions.

```{r fig-post-prob-test, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Posterior probability of the EEG difference (slope) being above 0 according to the GAM."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3, n_draws = 1e3) {

@@

```{r fig-post-prob-ratio, echo = FALSE, out.width = "100%", fig.width = 9, fig.asp = 0.5, fig.cap = "Left: Posterior probability of the EEG difference (slope) being above 0 according to the GAM. Right: Ratio of posterior probability according to the GAM (on a log10 scale). Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

## Multilevel modelling using summary statistics

The previous model only included constant (fixed) effects, thus not properly accounting for between-participant variability. We next fit a multilevel version of the GAM. Although it is possible to fit a GAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD), similar to what is done in meta-analysis.

@@

We then compute the difference between the true and estimated onset/offset of the EEG difference according to various `k` (GAM basis dimension) and `threshold` values. Remember that the EEG signal was generated from a truncated Gaussian with an objective onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms. @fig-onset-error shows that the multilevel GAM can *almost exactly* recover the true onset and offset values, given some reasonable choice of `k` and `threshold` values (e.g., a threshold of 20). We provide more detailed recommendations on how to set `k` in @sec-basis.

```{r fig-onset-error, eval = TRUE, echo = FALSE, results = "hide", dev = "png", dpi = 300, fig.width = 9, fig.asp = 0.5, fig.cap = "Average estimation error for the onset (left) and offset (right) according to various basis dimension and threshold values (according to the multilevel GAM)."}
# defining values of k and threshold to test
# threshold_values <- seq(from = 1, to = 50, by = 1)
# kvalues <- c(5, 10, 15, 20, 25, 30, 35, 40)

@@

# number of posterior samples to use
# n_post_samples <- 1e4

@@

# initialising results dataframe
# results <- crossing(kvalue = kvalues, sim_id = 1:nsims, threshold = threshold_values) %>%
#     mutate(
#         estimated_onset = NA,
#         estimated_offset = NA,
#         error_onset = NA,
#         error_offset = NA
#         )

# looping over different values of k and threshold
# for (i in seq_len(nrow(results) ) ) {
# 
#     # retrieving current kvalue and threshold
#     kvalue <- results$kvalue[i]
#     threshold <- results$threshold[i]
# 
#     # printing progress
#     cat(
#         "\nAssessing combination:", i, "out of", nrow(results), "combinations...\n",
#         "k-value:", kvalue, "\n", "threshold:", threshold, "\n"
#         )
# 
#     # when we assess a new k-value or new sim_id, we generate new data
#     # and fit a new model...
#     if (threshold == 1) {
# 
#         # generate some data
#         raw_df <- generate_data(
#             n_trials = n_trials, n_ppt = n_ppt, outvar = outvar,
#             srate = srate, ronset = ronset
#             )
# 
#         # sanity check
#         # head(raw_df)
# 
#         # averaging across participants
#         summary_df <- raw_df %>%
#             summarise(
#                 eeg_mean = mean(eeg),
#                 eeg_sd = sd(eeg),
#                 .by = c(participant, condition, time)
#                 )
# 
#         # defining a contrast for condition
#         contrasts(summary_df$condition) <- c(-0.5, 0.5)
# 
#         # fitting the GAM
#         error_model <- brm(
#             # using by-participant SD of EEG across trials
#             eeg_mean | se(eeg_sd) ~
#                 condition + s(time, bs = "cr", k = kvalue, by = condition) +
#                 (1 | participant),
#             data = summary_df,
#             family = gaussian(),
#             warmup = 2000,
#             iter = 5000,
#             chains = 8,
#             cores = 8,
#             ###################################
#             # when running on the HPC cluster #
#             ###################################
#             backend = "cmdstanr",
#             stan_model_args = list(stanc_options = list("O1") )
#             )
# 
#         prob_y_above_0 <- error_model$data %>%
#             # add_epred_draws(object = error_model, ndraws = n_post_samples) %>%
#             add_epred_draws(object = error_model) %>%
#             data.frame() %>%
#             dplyr::select(participant, time, condition, .epred, .draw) %>%
#             pivot_wider(names_from = condition, values_from = .epred) %>%
#             mutate(epred_diff = cond2 - cond1) %>%
#             # computing mean posterior probability at the group level
#             group_by(time) %>%
#             summarise(m = mean(epred_diff > 0) ) %>%
#             mutate(prob_ratio = m / (1 - m) ) %>%
#             ungroup() %>%
#             # ensuring there is no 0 or +Inf values
#             mutate(
#                 prob_ratio = ifelse(
#                     is.infinite(prob_ratio),
#                     ndraws(error_model),
#                     prob_ratio
#                     )
#                 ) %>%
#             mutate(
#                 prob_ratio = ifelse(
#                     prob_ratio == 0,
#                     1 / ndraws(error_model),
#                     prob_ratio
#                     )
#                 )
# 
#     }
# 
#     # finding onset, offset, and peak for the current threshold
#     exceeding_times <- prob_y_above_0 %>%
#         dplyr::filter(prob_ratio > threshold) %>%
#         summarise(
#             cluster_onset = min(time, na.rm = TRUE),
#             cluster_offset = max(time, na.rm = TRUE)
#             ) %>%
#         mutate(true_onset = true_onset, true_offset = true_offset)
# 
#     # storing the results in the dataframe
#     if (nrow(exceeding_times) > 0) {
# 
#         results$estimated_onset[i] <- exceeding_times$cluster_onset
#         results$estimated_offset[i] <- exceeding_times$cluster_offset
# 
#         # computing errors
#         results$error_onset[i] <- abs(
#             exceeding_times$cluster_onset - exceeding_times$true_onset
#             )
#         results$error_offset[i] <- abs(
#             exceeding_times$cluster_offset - exceeding_times$true_offset
#             )
# 
#     }
# 
# }

@@

# running the simulation on the HPC cluster
# source("code/meta_gam_errors_cluster.R")

# loading the simulation results
# results <- readRDS(file = "results/meta_gam_error_properties.rds")
results <- readRDS(file = "results/meta_gam_error_properties_cluster.rds")

# plotting the results
results %>%
    group_by(kvalue, threshold, sim_id) %>%
    # mutate(error_onset = ) %>%
    summarise(
        # value = if (onset_offset[1] == "onset") min(value) else max(value),
        cluster_onset = min(cluster_onset),
        cluster_offset = max(cluster_offset)
        # .groups = "drop"
        ) %>%
    ungroup() %>%
    group_by(kvalue, threshold) %>%
    # summarise(across(estimated_onset:error_offset, median) ) %>%
    # summarise(across(estimated_onset:error_offset, mean) ) %>%
    mutate(
        error_onset = abs(cluster_onset - true_onset),
        error_offset = abs(cluster_offset - true_offset)
        ) %>%
    summarise(across(error_onset:error_offset, mean) ) %>%
    ungroup() %>%
    pivot_longer(cols = error_onset:error_offset) %>%
    mutate(name = factor(x = name, levels = c("error_onset", "error_offset") ) ) %>%
    ggplot(
        aes(
            x = as.factor(kvalue),
            y = as.factor(threshold),
            fill = value
            )
        ) +
    geom_raster(interpolate = FALSE) +
    # geom_tile() +
    # geom_point(
    #     data = results %>%
    #         group_by(kvalue, threshold) %>%
    #         summarise(across(estimated_onset:error_offset, mean) ) %>%
    #         ungroup() %>%
    #         na.omit() %>%
    #         filter(error_onset == min(error_onset) ),
    #     # aes(x = kvalue, y = threshold),
    #     aes(x = kvalue, y = value),
    #     color = "orangered", size = 4, shape = 4,
    #     # show.legend = FALSE
    #     show.legend = TRUE
    #     ) +
    # scale_x_continuous(expand = c(0, 0) ) +
    # scale_y_continuous(expand = c(0, 0) ) +
    facet_wrap(~name) +
    scale_x_discrete(expand = c(0, 0) ) +
    scale_y_discrete(expand = c(0, 0), breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50) ) +
    # scale_y_continuous(n.breaks=10) +
    scale_fill_gradientn(colors = rev(met.brewer("Hokusai1") ) ) +
    # scale_fill_viridis_c(direction = 1) +
    labs(
        x = "GAM basis dimension (k)",
        y = "Threshold",
        fill = "Error (s)"
        )

@@

# displaying best parameters values (according to both onset and offset error)
# results %>%
#     group_by(kvalue, threshold) %>%
#     summarise(across(estimated_onset:error_offset, mean) ) %>%
#     # summarise(across(estimated_onset:error_offset, mode) ) %>%
#     ungroup() %>%
#     arrange(error_onset, error_offset) %>%
#     data.frame() %>%
#     head(40)
```

@@

We then compared the ability of the GAMM to correctly estimate the onset and offset of the ERP difference to other widely-used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.05$). We then followed the same approach but after applying different forms of multiplicity correction to the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the family-wise error rate (FWER) [i.e., Holm–Bonferroni method, @holm1979], and two cluster-based permutation methods [permutation with a single cluster-forming threshold and threshold-free cluster enhancement, TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. @fig-corrections illustrates the onsets and offsets estimated by each method on a single simulated dataset and shows that all methods over-estimate the true onset and under-estimate the true offset.

@@

```{r fig-corrections, echo = FALSE, results = "hide", fig.width = 8, fig.asp = 0.5, fig.cap = "Exemplary timecourse of squared t-values with true onset and offset (vertical black dashed lines) and onsets/offsets identified using the raw p-values, the corrected p-values (BH95, BY01, Holm), the cluster-based methods (Cluster mass, TFCE), or using the binary segmentation method (Change point)."}
# defining an arbitrary alpha threshold
aath <- 0.05

@@

The various methods were compared using the median error (i.e., true-estimated) and variance of onset/offset estimates computed on 10.000 simulated datasets.

@@

We assessed the reliability of the proposed approach using a form of permutation-based split-half reliability [as for instance in @rosenblatt2018], which consisted of the following steps. First, we created 1000 split halves of the data (i.e., with half the participants in the original data, that is, 16 participants). For each split, we estimated the onset/offset using all methods described previously. Third, we summarised the distribution of onset/offset estimates using the median "error" (i.e., difference between the split estimate and the estimate obtained using the full dataset) and the variance across splits. This approach allows assessing how similar the estimate of each half split is to the full dataset (thus acting as a proxy for the population) and how variable the estimates are (across split halves).

@@

## Simulation study (bias and variance)
